Omics-based Single Step Trait Prediction

### 2017-09-06 paper/manuscript.tex
paper, minor methods fixes

% Two-step prediction.
Initially, a two-step procedure was the most intuitive choice for accomplishing
this goal~\cite{VanRaden2009}.
The first step consists of conventionally estimating breeding values based on
pedigree relationships and the phenotypes of relatives.
Individuals with accurate breeding values out of the first step (that is a
high number of offspring), enter the second step in which conventional breeding
values are regressed onto SNP genotypes.
The estimated model coefficients can then be used to predict genomic breeding 
values for solely genotyped and potentially very young individuals.
However, as the breeding values from step one enter the genomic prediction
model as phenotypes, one has to be careful about the residual error
distribution of such a model.
For instance, the range of the prediction errors will no longer be identically
distributed, that is they should be weighted differently or estimated 
individually~\cite{Aguilar2010}.
One popular approach to account for this problem is deregressing the breeding
values and weighting the residuals according to the prediction error variances
\cite{Garrick2009}.

@@

\section{Materials and Methods}
\Genetics2level{Maize hybrids}
To address the objectives 1 and 2, a maize hybrid data set, subsequently
denoted as \textit{Exp1}, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines~\cite{Westhues2017}. 
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse locations across Germany for six 
agronomically important traits by accounting for year, location, field 
replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
using the Beagle software~\cite{Browning2009}.
The final number of polymorphic marker loci was 7,013 for the Dent and 6,212 for 
the Flint lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 parental Dent and 43 parental Flint lines, that are parents of 685
hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts while adjusting
for putative effects arising from the presence of multiple factorials
(\textit{cf.} \citeNP{Westhues2017}) using the R-package
\textit{limma}~\cite{Ritchie2015a}.

\Genetics2level{Maize diversity panel}
To address objectives 2 and 3 a maize inbred line data set, subsequently
referred to as \textit{Exp2}, was used~\cite{Yang2014}.
Exp2 originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$) given that it is the largest of the four pre-classified
subgroups~\cite{Yang2014}.
All inbred lines were evaluated in five Chinese environments ranging from
$18$ to $30^{\circ}$N and from $102$ to $110^{\circ}$E~\cite{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated for 17 traits of which 
six are presented in this study (Table~\ref{table:TraitKey}).
The selection of these traits was done so that four traits were better 
predicted by transcriptomic data and two further traits were better predicted 
by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploting identify by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines~\cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers~\cite{Yang2014}.
The same SNP quality checks as for the parental inbred lines of the hybrids
were applied to the 211 inbred lines from the tropical/subtropical subset,
yielding 37,760 SNPs.
Gene expression data were normalized using a normal quantile transformation to
satisfy modelling assumptions~\cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes and was kept for further analyses~\cite{Li2013}.

@@

\Genetics2level{Population structure}
Population structure in Exp2 was examined by running a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated probability of the data for each $K$, we chose $K=4$
and ran a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the R-pacakge LEA
\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \textit{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:inbred-pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To avoid inflated predictive abilities, which can arise from large differences
in the average performance of the different subpopulations
\cite{Windhausen2012}, we reduced the panel of inbred lines to those primarily
belonging to either cluster 1 or cluster 4 in Fig.~\ref{fig:inbred-pca}.

@@

where $\mathbf{y}$ is the vector of observed inbred line or hybrid performance,
respectively, $\mu$ is the fixed model intercept, and $\mathbf{e}$ is the 
residual error.
For the hybrid data from \textit{Exp1}, we fit separate additive effects for 
each of the $L$ heterotic groups and $\mathbf{g}_{l}$ now pertains to the 
general combining ability effects of the parental inbred lines (either Dent or 
Flint, as expressed by $l$).
For the maize diversity panel inbred lines from \textit{Exp2}, we fit random 
genotype effects $\mathbf{g}_{l}$ (\textit{i.e.}, breeding values) of the 
lines for the $o$-th predictor.
For both experiments, genetic effects $\mathbf{g}_{l}$ are associated with 
the vector of phenotypic performance $\mathbf{y}$ through a corresponding 
design matrix $\mathbf{Z}_{l}$.
The random effects ($\mathbf{g}_{l}$) have expectation zero and variance equal 
to $\mathbf{K}_{l} \sigma^{2}_{{g}^{l}}$ and $\mathbf{I} \sigma^2_{e}$ 
for the residual error.
Here, we did not consider the inclusion of dominance effects for the hybrid
material after a previous investigation into the predictive ability of various
``omics'' predictors in the same material concluded no considerably gains 
\cite{Westhues2017}.
All prediction models were implemented using the R-packages \textit{BGLR}
\cite{Perez2014} and \textit{sspredr}
\url{https://github.com/mwesthues/sspredr}.

@@

The structure of the residual imputation/prediction error is known and can 
therefore be modelled.
The covariates not covered by the incomplete predictor can be predicted using 
the expectation of a multivariate normal random vector given correlated 
observations, which are related to the Best Linear Predictor.

@@

\Genetics2level{Core sampling}
In order to evaluate the influence of the fraction of genotypes that is not
covered by the incomplete predictor on predictive ability we generated core
sets of nine different sizes from Exp2 (Fig.~\ref{fig:compound-plot}~a).
The base of each of these core sets was the superset of genotypes that were
covered by both, genomic and trancriptomic data.
A core set was created by artificially declaring a pre-specified fraction 
$x \in {10\%, 20\%, \dots, 90\%}$ of genotypes as lacking of transcriptomic 
information.
For each value of $x$ we created 100 core sets $s \in {1, 2, 3, \dots, 100}$ by
repeatedly sampling genotypes, whose trancsriptomic information was then 
declared as missing, at random (Fig.~\ref{fig:compound-plot}~b).
The artificially removed transriptomic information on these genotypes was then
imputed.

\begin{figure}[H]
  \centering
  \includegraphics[width=1\textwidth]{./tables_figures/compound_plot.pdf}
  \caption{
  (\textbf{a}) Scheme for the amount of genotypes covered by predictors.
  The two extended bars represent the entire set of genotypes in either Exp1 or
  Exp2.
  All genotypes are covered by at least one predictor, which is called the
  `complete' predictor and depicted as a white bar.
  The second bar depicts that only a fraction of genotypes is covered by both,
  the `complete' and the `incomplete' predictor (dark gray) and that the
  difference between the predictors will be imputed.
  The four bars to the right outline the principle of generating core sets for
  Exp2.
  For nine different values of $x$, which represents the fraction of genotypes
  covered by the `incomplete' predictor, core sets are build.
  (\textbf{b}) Subsampling within core sets.
  Each of the black symbols represents a different genotype.
  For each value of $x$, 100 subsets $s$ are generated where $x\%$ of genotypes
  are randomly declared to be covered by both, the `complete' and the
  `incomplete' predictor whereas the remaining $100\% - x\%$ of genotypes
  declared to be covered only by the `complete' predictor.
  }
\label{fig:compound-plot}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/pred_ability_inbred_core_fraction.pdf}
  \caption{
  Influence of the fraction of genotypes covered by the incomplete predictor
  on predictive abilities.
  Predictive abilities are based on transcriptomic effects derived from a
  single step prediction model involving transcriptomic (T) and genomic (G)
  data, respectively.
  Results are shown for the core set of 110 inbred lines from Exp2 for which
  both, G and T are available for any genotype and six agronomic traits.
  The fraction of genotypes (\textit{i.e.} `Core Fraction') for which
  T was available is given on the x-axis.
  Dotted lines indicate that $T > G$, dashed lines indicate that $T \approx G$
  and solid lines indicate that $G > T$ when running predictions using single
  predictors on the same data set.
  }
\label{fig:InbredResults}
\end{figure}

>>>>

% Two-step prediction.
Initially, a two-step procedure was the most intuitive choice for accomplishing
this goal~\cite{VanRaden2009}.
The first step consists of conventionally estimating breeding values based on
pedigree relationships and the phenotypes of relatives.
Individuals with accurate breeding values out of the first step (that is, a
high number of offspring), enter the second step in which conventional breeding
values are regressed onto SNP genotypes.
The estimated model coefficients can then be used to predict genomic breeding 
values for solely genotyped and potentially very young individuals.
However, as the breeding values from step one enter the genomic prediction
model as phenotypes, one has to be careful about the residual error
distribution of such a model.
For instance, the range of the prediction errors will no longer be identically
distributed, \textit{i.e.}, they should be weighted differently or estimated 
individually~\cite{Aguilar2010}.
One popular approach to account for this problem is deregressing the breeding
values and weighting the residuals according to the prediction error variances
\cite{Garrick2009}.

@@

\section{Materials and Methods}
\Genetics2level{Maize hybrids}
To address objectives 1 and 2, a maize hybrid data set, subsequently
denoted as \textit{Exp1}, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines~\cite{Westhues2017}. 
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse locations across Germany for six 
agronomically important traits by accounting for year, location, field 
replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
using the Beagle software~\cite{Browning2009}.
The final number of polymorphic marker loci was 7,013 for the Dent and 6,212 for 
the Flint lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 parental Dent and 43 parental Flint lines, that are parents of 685
hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts while adjusting
for putative effects arising from the presence of multiple factorials
(\textit{cf.} \citeNP{Westhues2017}) using the R-package
\textit{limma}~\cite{Ritchie2015a}.

\Genetics2level{Maize diversity panel}
To address objectives 2 and 3, a maize inbred line data set, subsequently
referred to as \textit{Exp2}, was used~\cite{Yang2014}.
\textit{Exp2} originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$) given that it is the largest of the four pre-classified
subgroups~\cite{Yang2014}.
All inbred lines were evaluated in five Chinese environments ranging from
$18$ to $30^{\circ}$N and from $102$ to $110^{\circ}$E~\cite{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated for 17 traits of which 
six are presented in this study (Table~\ref{table:TraitKey}).
The selection of these traits was done so that four traits were better 
predicted by transcriptomic data and two further traits were better predicted 
by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploting identify by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines~\cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers~\cite{Yang2014}.
The same SNP quality checks as for the parental inbred lines of the hybrids
were applied to the 211 inbred lines from the tropical/subtropical subset,
yielding 37,760 SNPs.
Gene expression data were normalized using a normal quantile transformation to
satisfy modelling assumptions~\cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes and was kept for further analyses~\cite{Li2013}.

@@

\Genetics2level{Population structure}
Population structure in \textit{Exp2} was examined by running a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated probability of the data for each $K$, we chose $K=4$
and ran a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the R-pacakge LEA
\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \textit{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:inbred-pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To avoid inflated predictive abilities, which can arise from large differences
in the average performance of the different subpopulations
\cite{Windhausen2012}, we reduced the panel of inbred lines to those primarily
belonging to either cluster 1 or cluster 4 in Fig.~\ref{fig:inbred-pca}.
This reduction yielded 164 genotypes, 110 of which were covered by both, genomic
and transcriptomic data.

@@

where $\mathbf{y}$ is the vector of observed inbred line or hybrid performance,
respectively, $\mu$ is the fixed model intercept, and $\mathbf{e}$ is the 
residual error.
For the hybrid data from \textit{Exp1}, we fit separate additive effects for 
each of the $L$ heterotic groups and $\mathbf{g}_{l}$ now pertains to the 
general combining ability effects of the parental inbred lines (either Dent or 
Flint, as expressed by $l$).
For the maize diversity panel inbred lines from \textit{Exp2}, we fit random 
genotype effects $\mathbf{g}_{l}$ (\textit{i.e.}, breeding values).
For both experiments, genetic effects $\mathbf{g}_{l}$ are associated with 
the vector of phenotypic performance $\mathbf{y}$ through a corresponding 
design matrix $\mathbf{Z}_{l}$.
The random effects ($\mathbf{g}_{l}$) have expectation zero and variance equal 
to $\mathbf{K}_{l} \sigma^{2}_{{g}^{l}}$ and $\mathbf{I} \sigma^2_{e}$ 
for the residual error.
Here, we did not consider the inclusion of dominance effects for the hybrid
material after a previous investigation into the predictive ability of various
``omics'' predictors in the same material concluded no considerably gains 
\cite{Westhues2017}.
All prediction models were implemented using the R-packages \textit{BGLR}
\cite{Perez2014} and \textit{sspredr}
\url{https://github.com/mwesthues/sspredr}.

@@

The structure of the residual imputation/prediction error is known and can 
therefore be modeled.
The covariates not covered by the incomplete predictor can be predicted using 
the expectation of a multivariate normal random vector given correlated 
observations, which are related to the Best Linear Predictor.

@@

\Genetics2level{Core sampling}
In order to evaluate the influence of the fraction of genotypes that is not
covered by the incomplete predictor on predictive ability we generated core
sets of nine different sizes from \textit{Exp2}
(Fig.~\ref{fig:compound-plot}~a).
The base of each of these core sets was the superset of genotypes that were
covered by both, genomic and trancriptomic data.
A core set was created by artificially declaring a pre-specified fraction 
$x = {10\%, 20\%, \dots, 90\%}$ of genotypes as lacking of transcriptomic 
information.
For each value of $x$ we created 100 core sets $s \in {1, 2, 3, \dots, 100}$ by
repeatedly sampling genotypes, whose trancsriptomic information was declared
as missing, at random (Fig.~\ref{fig:compound-plot}~b).
The artificially removed transriptomic information on these genotypes was then
imputed.

\begin{figure}[H]
  \centering
  \includegraphics[width=1\textwidth]{./tables_figures/compound_plot.pdf}
  \caption{
  (\textbf{a}) Scheme for the amount of genotypes covered by predictors.
  The two extended bars represent the entire set of genotypes in either
  \textit{Exp1} or \textit{Exp2}.
  All genotypes are covered by at least one predictor, which is called the
  `complete' predictor and depicted as a white bar.
  The second bar depicts that only a fraction of genotypes is covered by both,
  the `complete' and the `incomplete' predictor (dark gray) and that the
  difference between the predictors will be imputed.
  The four bars to the right outline the principle of generating core sets for
  \textit{Exp2}.
  For nine different values of $x$, which represents the fraction of genotypes
  covered by the `incomplete' predictor, core sets are build.
  (\textbf{b}) Subsampling within core sets.
  Each of the black symbols represents a different genotype.
  For each value of $x$, 100 subsets $s$ are generated where $x\%$ of genotypes
  are randomly declared to be covered by both, the `complete' and the
  `incomplete' predictor whereas the remaining $100\% - x\%$ of genotypes
  declared to be covered only by the `complete' predictor.
  }
\label{fig:compound-plot}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/pred_ability_inbred_core_fraction.pdf}
  \caption{
  Influence of the fraction of genotypes covered by the incomplete predictor
  on predictive abilities.
  Predictive abilities are based on transcriptomic effects derived from a
  single step prediction model involving transcriptomic (T) and genomic (G)
  data, respectively.
  Results are shown for the core set of 110 inbred lines from \textit{Exp2} for
  which both, G and T are available for any genotype and six agronomic traits.
  The fraction of genotypes (\textit{i.e.} `Core Fraction') for which
  T was available is given on the x-axis.
  Dotted lines indicate that $T > G$, dashed lines indicate that $T \approx G$
  and solid lines indicate that $G > T$ when running predictions using single
  predictors on the same data set.
  }
\label{fig:InbredResults}
\end{figure}
